### 2024-03-27 reviews/rebuttal.md

---
title: Rebuttal
author: Aliaume LOPEZ
date: 2024-03-27
lang: en-GB
---

We would like to thank the reviewers for their comments and suggestions, which
are all constructive and will definitely help to improve the quality of the
paper. In particular, we will add the reference to the ArXiv paper
`arXiv:2301.09234` in the introduction which is definitely relevant.

Before answering to the questions of the reviewers, we would like to point out
that the summary of Reviewer 1 is slightly misleading as the paper does not
manage to decide whether a function is "$\mathbb{N}$-polyregular" in general,
but only does so in the case where the function is assumed to be *commutative*
(which is a strong restriction). Without the assumption of commutativity, the
problem remains open. We will make sure to put more emphasis on this point.

---

It is our understanding that a significant part of the reviewers' concerns
relates to the high number of typos and grammatical errors in the manuscript.
We sincerely apologize for these mistakes that made reading this paper
difficult, and we acknowledge the lack of consistency in the bibliographic
section and in particular the mix of unicode and non unicode bibtex entries
which butchered the name of Lê Thành Dũng Tito Nguyên. We will definitely
correct these issues in the revised version of the manuscript. We decided to
not respond to specific typos and grammatical errors in this rebuttal and focus
on more global issues.

---

A second point raised by the reviewers is the lack of clarity for non-expert
readers, with a technical introduction which does not take enough time to
explain the state of the art, and for instance clearly state results obtained
in [9] to compare them with what is done in this paper. To address this issue,
we will take more time to introduce historical results and explain the various
classes of functions that have been studied before, expanding Figure 1 with
concrete examples. We will take time to introduce the various notions of
string-to-string functions (mealy machines, sequential transducers, rational
functions, regular functions, polyregular functions) with a clear table
comparing them and stating the status of "deciding aperiodicity" for each of
these classes.

However, it seems reasonable for us to assume that the reader is familiar at
least with finite monoids, monadic second order logic on words, and weighted
automata (hence, with rational series) which are well established theories:
describing them in a few sentences with a reference to a standard book should
already greatly increase the readability of the paper.

The informal uses of notions such as "growth rate" and "commutativity" in the
introduction were intended to provide intuitions to the reader, but we
understand that it was counterproductive and will be more careful in the
revised version. This is not an isolated issue, and we will make sure to avoid
such informal statements.

We also understand that whether a proof belongs to the appendix or the main
body of the paper felt arbitrary, and that the key ingredients of the proofs
where not clearly identified. While it is difficult to answer to such a comment
without rewriting the paper, we will try to illustrate on a few examples.

- Lemma 7 is a simple proof in the case of polyregular functions, and has
  a sligthly more technical proof[^1] in the case of rational series, but the
  main interest of the paper is in polyregular functions, and therefore the
  proof is in appendix.
- Lemma 19 is a technical result which while technically new, follows proof
  schemes similar to those of [9, 12], and provides another proof of the
  correct part of the theorem of [23]. This is why it is in the appendix.
- Lemma 26 is the actual core of Theorem 29, which is a corrected version of
  a theorem from [23]. In the (incorrect) proof of [23], the error lies in the
  inductive step of the analogue of Lemma 26, which is why we decided to put
  and this is why this particular proof was kept in the main body of the paper,
  while the other technical details about multivariate polynomials were moved
  to the appendix. 
- Lemma 31 has a proof that is a bit technical, but the important part of the
  lemma is actually the statement itself, which decomposes a commutative
  polyregular function into actual *polynomials*. Hence, the proof is in
  appendix.

However, we agree that all the proofs of sections 5 and 6 are in appendix
except for the ones that combine technical lemmas, and a better trade-off
between the sections will be found in the revised version of the paper. We
believe that the interesting proofs are those of Lemmas 44, and 45, which
really leverage the notion of well-quasi-ordering.

We will therefore revise the "Outlook" section to shape it more as a discussion
of the results and their implications, which will leave space for a more
detailed introduction, explanations of the results, and interesting proofs of
section 5.

---

To conclude on the relationship between this paper and [9], which should be
made clearer, let us try to replace both papers in context.

The concern of [9] is about $\mathbb{Z}$-polyregular functions (which are
a subclass of $\mathbb{Z}$-rational series), defining a notion of
"aperiodicity" for such functions, and proving that aperiodicity is decidable.
The work of [9] could be put into context as follows: since the introduction of
polyregular functions [4,5], there has been a corresponding notion of
"aperiodic" polyregular functions. However, decidability of aperiodicity
remains an open question, even in the specific case of unary output (the
function output words over a single letter alphabet). Notice that polyregular
functions with unary output are in fact functions from $\Sigma^*$ to
$\mathbb{N}$ (we write a number in unary).

While the work of [9] could be interpreted as studying $\mathbb{Z}$-rational
series, it is really about deciding aperiodicity of a "variation" of
polyregular functions with unary output that allows "negative outputs".
Unfortunately this "variation" is substantial, and the relationship between
$\mathbb{Z}$-polyregular functions (of [9]) and $\mathbb{N}$-polyregular
functions (unary output) remained unclear up to this paper. In particular the
decidability of aperiodicity for $\mathbb{N}$-polyregular functions could not
be deduced directly from [9].

This paper provides the combinatorial tools to understand the differences
between $\mathbb{Z}$-polyregular functions and $\mathbb{N}$-polyregular
functions (for instance, by giving decision procedures), and leverages these
tools to prove that aperiodicity is decidable for the latter **in the
commutative case**. This is a significant step towards the decidability of
aperiodicity for polyregular functions. The next step would be to drop the
assumption of commutativity, and then to allow the outputs to non-unary
alphabets. Both steps are beyond the scope of the current paper.

We hope that the above clarifications and proposed changes will address the
concerns of the reviewers, and would like to thank them again for the
constructive feedback.

[^1]: There is actually a simpler proof of Lemma 7 in the case of rational
series that avoids the use of Zariski closures and is more direct.

### 2024-03-28 parts/conclusion.tex
update ICALP
update bibliography

In the specific subclass of \emph{polyblind functions}
\cite{nguyn21comparison,doueneau2022hiding}, parts of the conjecture holds,
because they essentially behave as commutative functions \cite[Theorem
6.12]{gaetanphd}. Indeed, a suitable application of Parikh's Theorem
\cite{parikh66context} guarantees the existence of the functions $\mathsf{enc}$
and $\mathsf{dec}$, although they may fail to be  \kl{polyregular functions}.
We consider this partial result as a strong indicator in favor of the
conjecture.

>>>>

In the specific subclass of \emph{polyblind functions}
\cite{LENP21,DOUE22}, parts of the conjecture holds,
because they essentially behave as commutative functions \cite[Theorem
6.12]{DOUE23}. Indeed, a suitable application of Parikh's Theorem
\cite{PARI66} guarantees the existence of the functions $\mathsf{enc}$
and $\mathsf{dec}$, although they may fail to be  \kl{polyregular functions}.
Although \cref{z-poly-commutative-encoding:remark} is not yet applicable in
practice, we consider this partial result as a strong indicator in favor of the
conjecture.

### 2024-03-28 parts/intro.tex
update bibliography

\AP
Traditional results from automata theory state the equivalence of languages
defined in terms of deterministic finite automata and regular expressions
\cite{kleene1956representation}, Monadic Second Order Logic ($\intro*\MSO$)
\cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
Among regular languages, there exists a robust subclass sharing similar
characterizations, namely the \emph{star-free} regular languages. These are
obtained via counter-free (minimal) automata \cite{mcnaughton1971counter},
represented by star-free regular expressions \cite{schutzenberger1965finite},
definable in First Order Logic ($\intro*\FO$) \cite{perrin1986first}, and recognized
by aperiodic monoids \cite{schutzenberger1965finite}. It is decidable whether a
regular language is aperiodic, thanks to the characterizations in terms of
minimal automaton and/or syntactic monoid.

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness (for
instance, sequential, rational, regular, and \kl{polyregular functions})
\cite{EM65,Aho1969,eilenberg1974automata,EM02}. In this paper, we will focus on
\intro{polyregular functions} ($\intro*\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite{bojanczyk2019string}. As for regular languages, the class $\intro*\SF$ of
\intro{star-free polyregular function} can equivalently be introduced in terms
of first-order definability, counter-free pebble transducers, and suitable
restrictions of for-tranducers.

\AP
Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton. For
sequential functions (that admit a canonical object), aperiodicity is decidable
\cite{choffrut03}, and more recently, aperiodicity has been proven decidable
for rational functions \cite{filiot2016aperiodicity,filiot2018canonical}. The
problem remains open for regular functions, although it has been proven
decidable when changing the semantics of function equality to account for the
\emph{origins} in the output \cite{bojanczyk14}. To construct subclasses of
\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
of a function $f$, which is given by a number $k \in \Nat$, which can be used
to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
controls the maximal complexity of the $\MSO$ interpretation needed to
represent the function \cite{bojanczyk2022transducers}.

\AP
Another way to approach the problem of deciding if a \kl{polyregular function}
is \kl{star-free}, is to not restrict the computational power (via the
\kl{growth rate}), nor change the semantics (by adding origins), but consider
restrictions of the output alphabet. In particular, there have been successful
results proven on \emph{unary polyregular functions}, also known as
\kl{$\Nat$-polyregular functions} ($\intro*\NPoly$)
\cite{doueneau2021pebble,doueneau2022hiding}. For instance, the \kl{growth
rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
needed to compute $f$ \cite{doueneau2021pebble}, which does not hold on $\Poly$
\cite{bojanczyk2022transducers}. Let us remark that the restriction to
\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
outputs in the free commuative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{unary output} is really about
enforcing commutativity on the output.

\AP
Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). Although the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$).

\begin{conjecture}[{\cite[Conjecture 7.61]{gaetanphd}}]
    \label{zsf-nsf:conjecture}
    Let $k \in \Nat$.
    A function $f \in \NPoly[k]$
    belongs to $\NSF[k]$ if and only if
    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
    In particular,
    $\NSF[k] = \ZSF[k] \cap \NPoly$.
\end{conjecture}

\AP
It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
ambiguity, and a notion of \emph{quantitative} $\MSO$
\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
theorems obtained on these \cite{berstel2011noncommutative}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
\cite[Open question 5.55]{gaetanphd}.

@@

The approach taken in this paper avoids leveraging tools from the theory of
factorizations forests in finite monoids \cite{simon90}, which has been the
traditional method used to understand (commutative output) \kl{polyregular
functions} \cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, and instead
focus on intrinsic description of the functions themselves (via canonical
computational models such as the \kl{residual transducer} of
\cref{residual-transducer:def}, or explicit encoding/decoding into polynomials
via \cref{decompose-polynomial:lem}).

>>>>

\AP Traditional results from automata theory state the equivalence of languages
defined in terms of deterministic finite automata and regular expressions
\cite{KLEE56}, Monadic Second Order Logic ($\intro*\MSO$) \cite{BUCH60}, or
finite monoids \cite{SCHU61}. Among regular languages, there exists a robust
subclass sharing similar characterizations, namely the \emph{star-free} regular
languages. These are obtained via counter-free (minimal) automata
\cite{MNPA71}, represented by star-free regular expressions \cite{SCHU65},
definable in First Order Logic ($\intro*\FO$) \cite{PEPI86}, and recognized by
aperiodic monoids \cite{SCHU65}. It is decidable whether a regular language is
aperiodic, thanks to the characterizations in terms of minimal automaton and/or
syntactic monoid.

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness (for
instance, sequential, rational, regular, and \kl{polyregular functions})
\cite{ELME65,AHUL69,EILE74,ENMA02}. In this paper, we will focus on
\intro{polyregular functions} ($\intro*\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite{BOKL19}. As for regular languages, the class $\intro*\SF$ of
\intro{star-free polyregular function} can equivalently be introduced in terms
of first-order definability, counter-free pebble transducers, and suitable
restrictions of for-tranducers.

\AP
Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton. For
sequential functions (that admit a canonical object), aperiodicity is decidable
\cite{CHOF03}, and more recently, aperiodicity has been proven decidable
for rational functions \cite{FGL16,FGLM18}. The
problem remains open for regular functions, although it has been proven
decidable when changing the semantics of function equality to account for the
\emph{origins} in the output \cite{BOJA14}. To construct subclasses of
\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
of a function $f$, which is given by a number $k \in \Nat$, which can be used
to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
controls the maximal complexity of the $\MSO$ interpretation needed to
represent the function \cite{BOJA22}.

\AP
Another way to approach the problem of deciding if a \kl{polyregular function}
is \kl{star-free}, is to not restrict the computational power (via the
\kl{growth rate}), nor change the semantics (by adding origins), but consider
restrictions of the output alphabet. In particular, there have been successful
results proven on \emph{unary polyregular functions}, also known as
\kl{$\Nat$-polyregular functions} ($\intro*\NPoly$)
\cite{DOUE21,DOUE22}. For instance, the \kl{growth
rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
needed to compute $f$ \cite{DOUE21}, which does not hold on $\Poly$
\cite{BOJA22}. Let us remark that the restriction to
\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
outputs in the free commuative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{unary output} is really about
enforcing commutativity on the output.

\AP
Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
\cite{CDTL23}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). Although the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$).

\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
    \label{zsf-nsf:conjecture}
    Let $k \in \Nat$.
    A function $f \in \NPoly[k]$
    belongs to $\NSF[k]$ if and only if
    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
    In particular,
    $\NSF[k] = \ZSF[k] \cap \NPoly$.
\end{conjecture}

\AP
It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
ambiguity, and a notion of \emph{quantitative} $\MSO$
\cite{SCHU62,KRRC13}, and therefore can benefit from generic
theorems obtained on these \cite{BERE10}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{DRGA19,REUT80,CDTL23}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
\cite[Open question 5.55]{DOUE23}.

@@

The approach taken in this paper avoids the use of tools from the theory of
factorizations forests in finite monoids \cite{SIMO90}, which has been the
traditional method used to understand (commutative output) \kl{polyregular
functions} \cite{DOUE21,DOUE22,CDTL23}, and instead
focuses on intrinsic description of the functions themselves (via canonical
computational models such as the \kl{residual transducer} of
\cref{residual-transducer:def}, or explicit encoding/decoding into polynomials
via \cref{decompose-polynomial:lem}).

### 2024-03-28 pnrs.tex
update proofs and paper
writing to publish
lipics compiles
update
big big rewrite
update bibliography

%! TeX program = xelatex
%! lang = en-US
\documentclass[sigconf,natbib=false,screen, review,anonymous]{acmart}

@@

% CONFIG
\input{globals/macros}
\input{globals/mathdefs}
\input{globals/knowledges.kl}
\input{globals/metadata}

@@

\makeabstract
% DOCUMENT
\begin{document}

% FRONT MATTER
\maketitle
\acknowledge

\input{parts/intro.tex}
\input{parts/prelim.tex}
\input{parts/counter.tex}
\input{parts/correction.tex}
\input{parts/starfree.tex}
\input{parts/conclusion.tex}

% BACKMATTER
\printbibliography

@@

>>>>

%! TeX program = xelatex
%! TeX root = pnrs.tex
%! lang = en-US
\documentclass[anonymous,a4paper,UKenglish,cleveref,autoref,thm-restate]{lipics-v2021}

@@

% CONFIG
\input{globals/macros}
\input{globals/mathdefs}
\input{globals/knowledges.kl}

@@

\crefname{fact}{Fact}{Facts}
\Crefname{fact}{Fact}{Facts}
\crefname{algocf}{Algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}

% DOCUMENT
\begin{document}
\input{globals/metadata}

% FRONT MATTER
\maketitle
\makeabstract
%\acknowledge

\input{parts/intro.tex}
\input{parts/prelim.tex}
\input{parts/counter.tex}
\input{parts/correction.tex}
\input{parts/starfree.tex}
\input{parts/transducers.tex}
\input{parts/conclusion.tex}

% BACKMATTER
\bibliographystyle{plainurl}
\bibliography{transducer-bib/automata,%
transducer-bib/model-theory,%
transducer-bib/sublinear,%
transducer-bib/rational-series,%
transducer-bib/polyregular,%
transducer-bib/wqos}

@@

\include{parts/appendix.tex}
